From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-06 Pi+/paper/ufin.tex
intro to ufin

The denotational semantics for $\PiLang$ in the previous section forms a groupoid $\BFin$ whose objects are finite sets
and whose morphisms as bijections between finite sets. The semantics interprets types as objects in~$\BFin$,
1-combinators as isomorphisms in~$\BFin$, and for every pair of 1-combinators related by a 2-combinator, their
interpretations are extensionally equal in~$\BFin$. In other words, this groupoid is strict: two equal permutations are
extensionally identified, without an explicit witness for the equality. In order to establish completeness, we want to
quote back from the semantics to the syntax. Specifically, given a morphism in $\BFin$, we want to produce a
1-combinator in $\PiLang$, and given two extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator
witnessing the equality of the quoted 1-combinators.

Our first step is therefore to expose the implicit equalities by \emph{weakening} the groupoid $\BFin$, i.e., by
replacing the equational axioms of the category by higher invertible cells. In a weak category, the identity and
associativity equational conditions are replaced by higher invertible cells for the left and right units of composition
and for associativity; these higher cells come with their own coherence data as well, such as vertical composition, and
horizontal composition, or whiskering as detailed in Appendix~\ref{app:twocat}. However, although weakening the groupoid
exposes some implicit equalities, it does not, by itself, give us a handle to characterize such equalities. One elegant
way to acquire such an ability is to ensure that the groupoid is ``univalent,'' as this provides a way to characterize
the semantic equivalences using the elements of the identity type, i.e., using objects whose equality is equipped with
an induction principle. We will therefore work within a HoTT metatheory which provides an ambient univalent universe of
types and introduce the novel idea of a \emph{univalent subuniverse} to carve out $\UFin$: a weakened, univalent,
version of the groupoid $\BFin$.

>>>>

The denotational semantics for $\PiLang$ in the previous section forms a commutative rig groupoid $\BFin$ whose objects
are finite sets, whose morphisms as bijections between finite sets, and whose monoidal structures are induced by the
properties $\sqcup$ and $\times$. The semantics interprets types as objects in~$\BFin$, 1-combinators as isomorphisms
in~$\BFin$, and for every pair of 1-combinators related by a 2-combinator, their interpretations are extensionally equal
in~$\BFin$. In other words, this groupoid is strict: two equal permutations are extensionally identified, without an
explicit witness for the equality. In order to establish completeness, we want to quote back from the semantics to the
syntax. Specifically, given a morphism in $\BFin$, we want to produce a 1-combinator in $\PiLang$, and given two
extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator witnessing the equality of the quoted
1-combinators.

Our first step is therefore to expose the implicit equalities by \emph{weakening} the groupoid $\BFin$, i.e., by
replacing the equational axioms of the category by higher invertible cells. In a weak category, the identity and
associativity equational conditions are replaced by higher invertible cells for the left and right units of composition
and for associativity; these higher cells come with their own coherence data as well, such as vertical composition, and
horizontal composition, or whiskering as detailed in Appendix~\ref{app:twocat}. However, although weakening the groupoid
exposes some implicit equalities, it does not, by itself, give us a handle to characterize such equalities. One elegant
way to acquire such an ability is to ensure that the groupoid is ``univalent,'' as this provides a way to characterize
the semantic equivalences using the elements of the identity type, i.e., using objects whose equality is equipped with
an induction principle. We will therefore work within a HoTT metatheory which provides an ambient univalent universe of
types and introduce the novel idea of a \emph{univalent subuniverse} to carve out $\UFin$: a weakened, univalent,
version of the groupoid $\BFin$ which is then extended with the appropriate commutative rig structure. The main
technical result of this section that is exploited in the next section is that, for each finite type $T$, we have
transferred the structure of the space of equivalences $T \eqv T$ to the structure of paths in $\UFin$. Since the space
equivalences forms the automorphism group of $T$, we can now use the tools of computational group theory together with
the induction principle of paths and term rewriting theory to describe the paths in $\UFin$.
